 

### 2021-08-22 vertical-thermal-review.Rmd
more edits

**Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor, with the profile shape modified by LAD, canopy height, canopy structure, and across species and forest types [Figs. `r fig_NEON_vertical`b-d, `r SIfig_NEON_all`, @koike_leaf_2001].** 
This gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`d), than in open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Figs. `r fig_NEON_vertical`d, `r SIfig_NEON_all`, @aussenacInteractionsForestStands2000a; @smithSeasonalDroughtrelatedChanges2019; @tymenQuantifyingMicroenvironmentalVariation2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997]. 
Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].

**Wind speeds are also higher at the top of the canopy, owing to the buffering effect of the canopy [@juckerCanopyStructureTopography2018].**
Wind speeds increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018] to savannas [@curtis_intracanopy_2019; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018], including all 22 forested NEON sites (Figs. `r fig_NEON_vertical`e, `r SIfig_NEON_all`). Specifically, averaged across the NEON sites, maximum daily wind speeds were 4.1 m s^-1^ at the top of the vertical profile, compared to  $\le$ 1.1 m s^-1^ at the bottom (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).
The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_a$) and associated potential for both water and heat loss (see section "Leaf temperature").

@@

**Fundamentally, $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on biophysical principles, where $T_{leaf}-T_{air}$ is a function of energy input from net radiation ($R_n$, including shortwave and longwave) minus heat lost to the environment [Fig. `r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019].**
High $R_n$ loads can elevate $T_{leaf}$ dramatically above $T_{air}$ (Fig. `r fig_leaf_T`a).Sensible heat flux between leaf and air is regulated by leaf boundary layer conductance ($g_a$), which is greater in smaller leaves (Fig. `r fig_leaf_T`d) and higher wind speeds [Fig. `r fig_leaf_T`b]. Latent heat flux ($\lambda E$) through transpiration has a strong cooling effect, and is determined by stomatal and boundary layer conductances ($g_s$ and $g_b$) and VPD; notably, $g_b$ is influenced by windspeed and $g_s$ respond due to stomatal closure under high VPD.
Therefore, under full sunlight, $T_{leaf}-T_{air}$ decreases with $g_b$, increases with wind speed [Fig. `r fig_leaf_T`b,@daudetWindSpeedLeaf1999a]; decreases with leaf size (Fig. `r fig_leaf_T`d);
decreases with $g_s$ (Fig. `r fig_leaf_T`e); and increases with VPD, or increases with RH (Fig. `r fig_leaf_T`c).

@@

**Leaf energy balance is a tradeoff between water-use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].** 
When $g_s$ is high, high $\lambda E$ facilitates heat dissipation. 
However, when leaf water demand (influenced by VPD and photosynthetic CO~2~ demand) exceeds the rate of supply, conservation of water ($g_s$-limitation) occurs at the cost of increasing $T_{leaf}$ [Fig. `r fig_leaf_T`e, @fauset_differences_2018].
Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar radiation loads, but during $g_s$-limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021].
With adequate water, species with smaller leaves (higher $g_a$) and larger $g_s$ will be most effective at cooling under high temperatures [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts2021; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].

@@

**These basic biophysical principles shape the propensity for $T_{leaf}$ to be elevated above $T_{air}$ across vertical gradients in closed-canopy forests.**
High radiation at the top of the vertical profile (Fig. `r fig_NEON_vertical`) implies that canopy leaves have much greater propensity for high $T_{leaf}-T_{air}$, such that they would be expected to be warmer than understory leaves under most conditions (Fig. `r fig_leaf_T`).
However, higher wind speeds (e.g., up to ~ 4 $ms^{-1}$ at NEON sites; Fig. `r fig_NEON_vertical`) are capable of lowering $T_{leaf}-T_{air}$ upto ~ $4^\circ$C [Fig. `r fig_leaf_T`b, @bonan_ecological_2016; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. Collectively, higher wind speed (Fig. `r fig_NEON_vertical`), and with adaptive leaf traits increase $g_a$ and $g_s$ (reviewed below) mediate the direct effects of solar radiation on upper canopy leaves (Fig. `r fig_leaf_T`a) and result in higher rates of heat loss (Fig. `r fig_leaf_T`b-e). In contrast, greater RH and lower wind speeds in the understory may stifle $\lambda E$ (through reduced $g_s$ and $g_b$), thereby increasing $T_{leaf}$ [Fig. `r fig_leaf_T`, @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979b].

@@

**Within closed-canopy forests vertical gradients, the propensity for $T_{leaf}$ to be elevated above $T_{air}$ varies with micrometeorological conditions (Fig. `r fig_leaf_T`).**
Under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a propensity for $T_{leaf} - T_{air}$ to be elevated in the canopy compared to the understory (Fig.`r fig_leaf_T`).
Because sun leaves are exposed to higher irradiance and VPD, their $g_s$-limitation is steeper than shade leaves, resulting in greater increases in $T_{leaf}$ when $g_s$ is reduced [Fig. `r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018]. 
In contrast, under hot and wet conditions, higher wind speeds and $g_a$ enable cooling. Under lower wind conditions, the thickness of the air layer at the leaf surface curbs rapid leaf-to-air heat transfer, limiting $g_a$ and increasing $T_{leaf} - T_{air}$ [Fig. `r fig_leaf_T`b, @martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @songTreeSurfaceTemperature2020]. 
Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is not conducive to shedding excess heat, which in combination with their physiological propensities can result in greater $T_{leaf}$ under slight heat stress [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].
 In addition to lower wind speeds (Fig. `r fig_NEON_vertical`), higher RH in the understory (Fig. `r fig_NEON_vertical`) would also inhibit cooling, as $\lambda E$ is stifled under high RH, resulting in  greater $T_{leaf}$-$T_{air}$ [Fig. `r fig_leaf_T`c, @songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts2021].

@@

**Following the principles outlined above, vertical microclimate profiles, leaf traits, and $g_s$ together explain *in situ* $T_{leaf}$ observations.** 
Often, $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s$,* and therefore differs little between understory and canopy.
In forests with dense canopies, including tropical and temperate broadleaf forests, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopy than buffered lower-canopies. In some cases canopy leaves exceed the optima for photosynthesis ($T_{opt}$) and $T_{air}$ because of $g_s$-limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018]. 
In contrast, in open forests with lower LAI, lower-canopy $T_{leaf}$ can equal or even exceed upper canopy $T_{leaf}$, due to greater light permeability, vertical air mixing, and still air in the lower canopy and heating from the ground [Figs. `r fig_NEON_vertical`h, `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987]. 
Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020 (in review)]. 
Greater $RH$ in the inner canopy can also increase $T_{leaf}$ relative to upper canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].

@@

**Leaf traits determine the capacity of leaves to thermoregulate [@michaletzEnergeticCarbonEconomic2016], and as such are critical to optimizing leaf metabolism across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004].**
Leaf morphology adjusts to the local microenvironment during leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of direct to diffused light and associated water demands determines the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016]. 
This dichotomy in leaves is observed along the vertical gradient where canopy-leaves vary dramatically from canopy-interior and understory due to the differences in biophysical conditions (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).

Across the vertical gradient, traits vary (1) across sun and shade leaves within individuals, (2) across canopy and understory individuals of the same species, and (3) across canopy and understory species. 
The majority of studies characterizing variation in leaf traits or metabolism examine intraspecific patterns (categories 1 and 2), which are therefore the main focus of this review (Table 1).
However, when it comes to understanding and modeling forest ecosystem function, given that species traits strongly shape the propensity for trees to inhabit understory vs. canopy positions, ecosystem function is also shaped by interspecific trait variation (e.g., *Lamour, Serbin in prep*).

@@

Within-canopy, leaf traits vary along the vertical light gradient [*i.e.,* sun and shade leaves, @casasSunShadeLeaves2011] at an anatomical, structural and biochemical level [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020]. 
As much of our understanding of trait coordination is based on the leaf economics spectrum concept [@wrightWorldwideLeafEconomics2004] which was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], further research is needed to characterize trait relationships and responses vertically through the canopy. 
It has been established that independent of the light environment, traits also vary within species along the ontogenetic trajectory from the understory to the canopy. For example, leaf mass per area (LMA), photosynthetic capacity ($A_A$) and $g_s$ have been observed to increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000]. Therefore, ontogeny is bound to play a role in thermal sensitivity [@carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a].

@@

Sun leaves have traits that reduce heat and water stress -- for instance, by being smaller, hairier, having higher vein density [Table 1, @leighInfluenceLeafSize2017; @vogelSunLeavesShade1968a; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a], and a steeper leaf-angle [@niinemetsAdjustmentFoliageStructure1998, *REF*]. These traits collectively aid heat dissipation and regulate $T_{leaf}$-photosynthetic balance. In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA), but they are more likely to overheat than sun leaves due to greater thermal capacitance [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b]. In open canopies, where light is comparatively homogeneous, exposed leaves are primarily shaped by $T_{air_{max}}$ and VPD stress, rather than solely by light [@mediavillaFoliarPlasticityRelated2019a].

@@

Elemental traits, such as $N_{area}$ and $P_{area}$, contribute to activities of high leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves [Table 1, @scartazzaInvestigatingEuropeanBeech2016; @niinemetsWithincanopyVariationRate2004a; @weerasingheCanopyPositionAffects2014]. Within-canopy leaf nitrogen distribution can be regulated to prevent photoinhibition in shade leaves during sunflecks, and direct light for sun leaves [@kitaoCanopyNitrogenDistribution2018]. Therefore, we would expect elemental traits to influence vertical foliar thermal sensitivity;  however, @zhuPlasticityPhotosyntheticHeat2018 and @osullivanThermalLimitsLeaf2017 observed no relation between foliar thermal tolerance and total foliar concentrations of N and P.

@@

The vertical structure of microenvironments likely influences forest photosynthetic capacity [@rey-sanchez_spatial_2016] via leaf phenology. For example, deciduousness is an adaptation to conserve water during dry periods [@meakemRoleTreeSize2018], but is also structured across the vertical profile. Within tropical deciduous species, the fraction of seasonally deciduous individuals increases toward larger tree size classes in hotter, brighter environments, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c]. Even in evergreen forests, environmental variation alters leaf age distributions [@albertCrypticPhenologyPlants2019], which affects forest photosynthetic capacity [@wuLeafDevelopmentDemography2016; @niinemetsLeafAgeDependent2016] and heat stress via age-specific leaf function [@albertCrypticPhenologyPlants2019]. For instance, in tropical evergreen forests, because leaf turnover rates of sun leaves in the upper canopy is faster than in the subcanopy, less damage accumulates [@millerOnlySunlitLeaves2021]. Leaf age affects vulnerability to heat stress and has a strong effect on metabolism  [@mariasImpactsLeafAge2017; @zhouLeafageEffectsTemperature2015; @kikuzawaSynthesisRelationshipsLeaf2006], and can be expected to similarly scale through vertically structured phenology to affect forest responses to heat anomalies.

@@

Vertical gradients in the biophysical environment shape which plant strategies, or sets of traits, are competitive in canopy versus understory conditions, thereby shaping community composition across size classes. 
Generally, the pattern of fundamental trait adaptations to light gradients across forest species is similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012]. 
For instance, across increasing tree size classes, there is an increase in the fractions of species that are deciduous [@conditQuantifyingDeciduousnessTropical2000c; @meakemRoleTreeSize2018].
However, the range in trait values can be several times greater in multi-species canopy than in understory species, based on their inherent species-specific traits and responses to environmental conditions. 
Therefore, vertical gradients in $T_{leaf}$ and metabolism within forest ecosystems are underpinned by both intra- and inter-specific variation in traits.

>>>>

**Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor, with the profile shape modified by LAD, canopy height, canopy structure, and across species and forest types [Figs. `r fig_NEON_vertical`a-d, `r SIfig_NEON_all`, @koike_leaf_2001].** 
This gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`d), than in open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Figs. `r fig_NEON_vertical`d, `r SIfig_NEON_all`, @aussenacInteractionsForestStands2000a; @smithSeasonalDroughtrelatedChanges2019; @tymenQuantifyingMicroenvironmentalVariation2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997]. 
Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].

**Wind speeds are also higher at the top of the canopy, owing to the buffering effect of the canopy [@juckerCanopyStructureTopography2018].**
Wind speeds increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018] to savannas [@curtis_intracanopy_2019; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018], including all 22 forested NEON sites (Figs. `r fig_NEON_vertical`e, `r SIfig_NEON_all`). Specifically, averaged across the NEON sites, maximum daily wind speeds were 4.1 m s^-1^ at the top of the vertical profile, compared to  $\le$ 1.1 m s^-1^ at the bottom (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).
The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_a$) and therefore greater potential for both water and heat loss (see section "Leaf temperature").

@@

**Fundamentally, $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on biophysical principles, where $T_{leaf}-T_{air}$ is a function of energy input from net radiation ($R_n$, including shortwave and longwave) minus heat lost to the environment [Fig. `r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019].**
High $R_n$ loads can elevate $T_{leaf}$ dramatically above $T_{air}$ (Fig. `r fig_leaf_T`a). Sensible heat flux between leaf and air is regulated by leaf boundary layer conductance, which is greater in smaller leaves (Fig. `r fig_leaf_T`d) and higher wind speeds (Fig. `r fig_leaf_T`b). Latent heat flux ($\lambda E$) through transpiration has a strong cooling effect, and is determined by stomatal and boundary layer conductances ($g_s$ and $g_b$) and VPD; notably, $g_b$ is influenced by windspeed and $g_s$ respond due to stomatal closure under high VPD.
Therefore, under full sunlight, $T_{leaf}-T_{air}$ decreases with $g_b$, which increases with wind speed [Fig. `r fig_leaf_T`b, @daudetWindSpeedLeaf1999a] and $g_s$ (Fig. `r fig_leaf_T`e); increases with leaf size (Fig. `r fig_leaf_T`d);
decreases with and increases with VPD, or increases with RH (Fig. `r fig_leaf_T`c).

@@

**Leaf energy balance is a tradeoff between water-use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].** 
When $g_s$ is high, high $\lambda E$ facilitates heat dissipation. 
However, when leaf water demand (influenced by VPD and stomatal opening) exceeds the rate of supply, conservation of water ($g_s$-limitation) occurs at the cost of increasing $T_{leaf}$ [Fig. `r fig_leaf_T`e, @fauset_differences_2018].
Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar radiation loads, but during $g_s$-limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021].
With adequate water, species with smaller leaves (higher $g_b$) and larger $g_s$ will be most effective at cooling under high temperatures [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts2021; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].

@@

**These basic biophysical principles shape the propensity for $T_{leaf}$ to be elevated above $T_{air}$ across vertical gradients in closed-canopy forests.**
High radiation at the top of the vertical profile (Fig. `r fig_NEON_vertical`) implies that canopy leaves have much greater propensity for high $T_{leaf}-T_{air}$, such that they would be expected to be warmer than understory leaves under most conditions (Fig. `r fig_leaf_T`).
However, higher wind speeds (e.g., up to ~ 4 $ms^{-1}$ at NEON sites; Fig. `r fig_NEON_vertical`) are capable of lowering $T_{leaf}-T_{air}$ upto ~ $4^\circ$C [Fig. `r fig_leaf_T`b, @bonan_ecological_2016; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. Collectively, higher wind speed (Fig. `r fig_NEON_vertical`), and with adaptive leaf traits increase $g_b$ and $g_s$ (reviewed below) mediate the direct effects of solar radiation on upper canopy leaves (Fig. `r fig_leaf_T`a) and result in higher rates of heat loss (Fig. `r fig_leaf_T`b-e). In contrast, greater RH and lower wind speeds in the understory may stifle $\lambda E$ (through reduced $g_s$ and $g_b$), thereby increasing $T_{leaf}$ [Fig. `r fig_leaf_T`, @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979b].

@@

**Within closed-canopy forests vertical gradients, the propensity for $T_{leaf}$ to be elevated above $T_{air}$ varies with micrometeorological conditions (Fig. `r fig_leaf_T`).**
Under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a propensity for $T_{leaf} - T_{air}$ to be elevated in the canopy compared to the understory (Fig.`r fig_leaf_T`).
Because sun leaves are exposed to higher irradiance and VPD, their $g_s$-limitation is steeper than shade leaves, resulting in greater increases in $T_{leaf}$ when $g_s$ is reduced [Fig. `r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018]. 
In contrast, under hot and wet conditions, higher wind speeds and $g_a$ enable cooling. Under lower wind conditions, the thickness of the air layer at the leaf surface curbs rapid leaf-to-air heat transfer, limiting $g_b$ and increasing $T_{leaf} - T_{air}$ [Fig. `r fig_leaf_T`b, @martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @songTreeSurfaceTemperature2020]. 
Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is not conducive to shedding excess heat, which in combination with their physiological propensities can result in greater $T_{leaf}$ under slight heat stress [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].
 In addition to lower wind speeds (Fig. `r fig_NEON_vertical`), higher RH in the understory (Fig. `r fig_NEON_vertical`) would also inhibit cooling, as $\lambda E$ is stifled under high RH, resulting in  greater $T_{leaf}$-$T_{air}$ [Fig. `r fig_leaf_T`c, @songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts2021].

@@

**Following the principles outlined above, vertical microclimate profiles, leaf traits, and $g_s$ together explain *in situ* $T_{leaf}$ observations.** 
Often, $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s$,* and therefore differs little between understory and canopy.
In forests with dense canopies, including tropical and temperate broadleaf forests, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopy than buffered lower-canopies. In some cases canopy leaves exceed the optima for photosynthesis ($T_{opt}$) and $T_{air}$ because of $g_s$-limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018]. 
In contrast, in open forests with lower LAI, lower-canopy $T_{leaf}$ can equal or even exceed upper canopy $T_{leaf}$, due to greater light transmission, vertical air mixing, and still air in the lower canopy and heating from the ground [Figs. `r fig_NEON_vertical`h, `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987]. 
Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020 (in review)]. 
Greater $RH$ in the inner canopy can also increase $T_{leaf}$ relative to upper canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].

@@

**Leaf traits determine the capacity of leaves to thermoregulate [@michaletzEnergeticCarbonEconomic2016], and as such are critical to optimizing leaf metabolism across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004].**
Leaf morphology adjusts to the previously experienced microenvironment during leaf formation in buds, and also during ongoing leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of far red to red light light and associated water demands determines the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016]. 
This dichotomy in leaves is observed along the vertical gradient where canopy-leaves vary dramatically from canopy-interior and understory due to the differences in biophysical conditions (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).

Across the vertical gradient, traits vary (1) across sun and shade leaves within individuals, (2) across canopy and understory individuals of the same species, and (3) across canopy and understory species. 
The majority of studies characterizing variation in leaf traits or metabolism examine intraspecific patterns (categories 1 and 2), which are therefore the main focus of this review (Table 1).
However, when it comes to understanding and modeling forest ecosystem function, given that species traits strongly shape the propensity for trees to inhabit understory vs. canopy positions, ecosystem function is also shaped by interspecific trait variation [e.g., *Lamour, Serbin in prep*; @cavaleri_height_2010].

@@

Within-canopy, leaf traits vary along the vertical light gradient [*i.e.,* sun and shade leaves, @casasSunShadeLeaves2011] at a biochemical, anatomical and structural level [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020]. 
As much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species [@wrightWorldwideLeafEconomics2004] which was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments. 
It has been established that independently of the light environment, traits also vary within species along the ontogenetic trajectory from the understory to the canopy. For example, leaf mass per area (LMA), photosynthetic capacity ($A_A$) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000]. Therefore, ontogeny is bound to play a role in thermal sensitivity [@carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a].

@@

Sun leaves have traits that reduce heat and water stress -- for instance, by being smaller, hairier, having higher vein density [Table 1, @leighInfluenceLeafSize2017; @vogelSunLeavesShade1968a; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a], and a steeper leaf-angle [@niinemetsAdjustmentFoliageStructure1998, *REF*]. These traits collectively aid heat dissipation and regulate $T_{leaf}$-photosynthetic balance. In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA), but they are more likely to overheat than sun leaves due to greater thermal capacitance [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b]. In open canopies, where light is comparatively homogeneous, exposed leaves are shaped more by $T_{air_{max}}$ and VPD stress than by light [@mediavillaFoliarPlasticityRelated2019a].

@@

Elemental concentrations such as $N_{area}$ and $P_{area}$, contribute to high rates of leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves [Table 1, @scartazzaInvestigatingEuropeanBeech2016; @niinemetsWithincanopyVariationRate2004a; @weerasingheCanopyPositionAffects2014], and may be optimized throughout the vertical canopy gradient [@buckleyOptimalCarbonPartitioning2021]. Within-canopy leaf nitrogen distribution can be regulated to prevent photoinhibition in shade leaves during sunflecks, and direct light for sun leaves [@kitaoCanopyNitrogenDistribution2018]. However, @zhuPlasticityPhotosyntheticHeat2018 and @osullivanThermalLimitsLeaf2017 observed no relation between total foliar concentrations of N and P and foliar thermal tolerance.

@@

The vertical structure of microenvironments likely influences forest photosynthetic capacity [@rey-sanchez_spatial_2016] via leaf phenology. For example, deciduousness is an adaptation to conserve water during cold or dry periods [@meakemRoleTreeSize2018], but is also structured across the vertical profile. Within tropical deciduous species, the fraction of seasonally deciduous individuals increases toward larger tree size classes in hotter, brighter environments, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c]. Even in evergreen forests, environmental variation alters leaf age distributions [@albertCrypticPhenologyPlants2019], which affects forest photosynthetic capacity [@wuLeafDevelopmentDemography2016; @niinemetsLeafAgeDependent2016] and heat stress via age-specific leaf function [@albertCrypticPhenologyPlants2019]. For instance, in tropical evergreen forests, because leaf turnover rates of sun leaves in the upper canopy is faster than in the subcanopy, less damage accumulates [@millerOnlySunlitLeaves2021]. Leaf age affects vulnerability to heat stress and has a strong effect on metabolism  [@mariasImpactsLeafAge2017; @zhouLeafageEffectsTemperature2015; @kikuzawaSynthesisRelationshipsLeaf2006], and can be expected to similarly scale through vertically structured phenology to affect forest responses to heat anomalies.

@@

Vertical gradients in the biophysical environment shape which plant strategies, or sets of traits, are competitive in canopy versus understory conditions, thereby shaping community composition across size classes. 
Generally, the pattern of fundamental trait adaptations to light gradients across forest species is similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012]. 
For instance, across increasing tree size classes, there is an increase in the fractions of species that are deciduous [@conditQuantifyingDeciduousnessTropical2000c; @meakemRoleTreeSize2018].However, exceptions exist: at least among evergreen species, shade tolerant species tend to have higher LMA than light demanding species, by contrast with the generally higher LMA for sun than shade leaves within canopies [@luskWhyAreEvergreen2008].
Further, the range in trait values can be several times greater in multi-species canopy than in understory species, based on their inherent species-specific traits and responses to environmental conditions. 
Therefore, vertical gradients in $T_{leaf}$ and metabolism within forest ecosystems are underpinned by both intra- and inter-specific variation in traits.
